Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-10 write-ups/PLDI/paper.tex
added cooperative example of graph traversal and some text. Also added quote by HSA

\paragraph{OpenCL concepts}
An OpenCL program is typically executed by many \emph{threads} running
in parallel. Threads are organized into \emph{workgroups} of equal
size. The memory is organized in three levels: \emph{global} memory is
accessible by any thread, \emph{local} memory is shared at the workgroup
level (i.e., threads of different workgroups cannot access the same
local memory), and \emph{private} memory is dedicated to thread-level
storage and cannot be shared between threads. These concepts are
illustrated over the two examples.

\paragraph{Work stealing}
Work stealing is generally used for computational work that can be
splitted into independently processable tasks. The processing of a given
task may create an arbitrary number of new tasks in such a way that it
is typically hard to know the total number of tasks in advance. Work
stealing enables the dynamic balance of task processing over several
processing units, OpenCL workgroups in our case. Each workgroup has an
associated task pool from which it obtains tasks to process and to where
it stores newly created tasks. When a workgroup's pool becomes empty, a
workgroup---rather than terminating---tries to \emph{steal} a task from
an other pool. Hence, workgroups keep on processing tasks until all the
work is done.

\myfiglong\ref{fig:workstealing} illustrates an OpenCL implementation of
a work stealing application. All threads start their execution at the
function marked with the $\mathsf{kernel}$ keyword. They receive a
pointer to the task pools that reside in global memory, and which were
initialized by the host program beforehand to contain the initial
tasks. Each thread can identify the workgroup it belongs to via the
function \getgroupid, and OpenCL provides a series of primitives in the
same vein to access for instance the thread global id, or its local id
within a workgroup. Here the group id is stored into the variable
$\mathsf(pool\_id)$ (which is, by default, in private memory) and used
to access the relevant task pool. The $\mathsf(pop\_or\_steal)$ function
is called by all threads of a workgroup, although inside only the thread
with a local id equal to $0$, also called \emph{master thread} of the
workgroup, will try to pop or steal a task from the pools. The master
thread uses local memory inside $\mathsf(pop\_or\_steal)$ to ensure that
the returned value is valid for all threads of its workgroup. If a task
was obtained, then all threads of the workgroup can participate in the
processing of the task inside the $\mathsf{process\_task}$
function. \HEComment{Manage to say that a task is considered big enough
to benefit from parallel computation} If some new tasks are created
during the processing, it is again the master thread who will be in
charge of pushing them to the pool. Although not depicted here,
concurrent accesses to each pool is guarded by a mutex which is
implemented by atomic compare and swap operations on global
memory. Henceforth, master threads may be spinning on mutexes, and
workgroups are also spinning on the main loop, so a fair scheduling of
workgroups is desirable.
\begin{figure}

@@

\caption{An excerpt of a work stealing algorithm in OpenCL}\label{fig:workstealing}
\end{figure}

>>>>

\begin{figure}
\ADComment{Tyler/Hugues: please insert workstealing example.}

\HEComment{standard version: one task pool per workgroup, the number of
workgroup is a constant. So we access our pool by pool\_id ==
get\_group\_id(0). When we try to steal, we can look into other pools
which are indexed by (pool\_id + delta) modulo get\_num\_workgroups(0).
}

@@

\HEComment{ cooperative version: add kill/fork, and must pass the number
of task pools explicitely since get\_num\_workgroups() cannot be used to
know the number of pools. }

\begin{lstlisting}
__kernel connect_four(__global uchar *base_board,
                      const int lookahead,
                      __global atomic_int *best_move,
                      __global int *all_done,
                      __global Node *tree,
                      __global Task *pools,
                      __global Mutex *pool_m,
                      const int num_pools)
{
  // init
  if (get_global_id(0) == 0) {
    init_first_tasks(tree, pools, num_pools);
  }
  global_barrier();

  // main loop
  while (true) {
    offer_kill();
    offer_fork();

    // obtain a task, or terminate if all work is done
    int pool_id = get_group_id(0) % num_pools;
    Task *t = pop_or_steal(pools, pool_m, pool_id, num_pools);
    if (t == NULL) {
      if (*all_done) { return; } else { continue; }
    }

    // perform task, may create new tasks
    Node *node = tree[t->node_index];
    value = node_value(base_board, node);
    if (winning(value) || node->level == lookahead) {
      // we reached a leaf
      update_parent_value(node, tree, best_move, all_done);
    } else {
      // create_children() pushes or donates new tasks
      create_children(node, tree, pools, pool_m, pool_id, num_pools);
    }
  }
}
\end{lstlisting}
\caption{An excerpt of a work stealing algorithm in OpenCL}\label{fig:workstealing}
\end{figure}
